\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-23 manuscript/main.tex
man: pls score

	\subsection{Preliminary step: Finite-basis Approximation }\label{sec: finite basis approximation}	
	Let ${b_m(t)}$ be a twice-differentiable basis of $L^2([0,1])$ whose second derivatives are also linearly independent, for example, cubic B-splines, the Fourier basis, or an orthonormal polynomial basis of degree greater than three. Using this basis, the $j$th functional predictor, regression coefficient,  PLS direction (suppressing iteration indices) are represented as follows:
	\begin{equation*}
		X_{ij}(t) = \sum_{m=1}^\infty \theta_{ijm} \, b_m(t),\quad
		\beta_j(t) = \sum_{m=1}^\infty \eta_{jm} \, b_m(t),\quad
		%%%%%%%%%%%%%%%%%%%%%%
		\xi_j(t)
		=
		\sum_{m=1}^\infty
		\gamma_{jm}
		b_m(t)
	\end{equation*}	
	In practice, the full set of coefficients can not be obtained  with finite sample size, as functional data are measured on a finite grid. Thus, we truncate the expansion at $M$ terms and define the truncated basis space
	$
	\mathcal{B} := \operatorname{span}\bigl(b_1(t), \ldots, b_M(t)\bigr).
	$
	We choose a moderately large $M$ (e.g., 15 or 20) to capture functional variation without fine-tuning, as smoothness is handled via penalization (see Section~\ref{sec: Regularization of PLS Components}).
	The truncated expansions of the predictor and coefficient are denoted as
	\begin{equation*}
		\widetilde{X}_{ij}(t) := \sum_{m=1}^M \theta_{ijm} \, b_m(t), \quad
		\tilde{\beta}_j(t) := \sum_{m=1}^M \eta_{jm} \, b_m(t),
	\end{equation*}
	and our suggest method restricts each PLS direction to admit the following expansion:
	\begin{equation*}
		\xi_j(t) = \sum_{m=1}^M \gamma_{jm} \, b_m(t).
	\end{equation*}
	This implies that all computations in this paper are carried out entirely within the subspace
	\begin{equation}\label{def:truncated_hilbert_space}
		\widetilde{ \mathcal{H} } := \mathcal{B}^K \times \mathbb{R}^p \subset \mathcal{H}.
	\end{equation}
	The $i$th hybrid predictor, projected on $\widetilde{ \mathcal{H} }$, is represented by the tuple
	\begin{equation*}
		\widetilde{W}_i := 
		(
		\widetilde{X}_{i1}, \ldots, \widetilde{X}_{iK}, \boldsymbol{Z}_i
		).
	\end{equation*}
	Let $\boldsymbol{\theta}_{ij}$, $\boldsymbol{\eta}_j$, and $\boldsymbol{\gamma}_j$ denote the $M$-dimensional vectors of coefficients:
	\begin{equation}\label{def:coef_vector_form}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top,
		\quad
		\boldsymbol{\eta}_{j} := (\eta_{j1}, \ldots, \eta_{jM})^\top ,
		\boldsymbol{\gamma}_{j} := (\gamma_{j1}, \ldots, \gamma_{jM})^\top.
	\end{equation}
	For the predictors, we stack the coefficient vectors across observations into the matrix
	\begin{equation}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	\end{equation}
	The full data for the hybrid PLS problem can be represented by the tuple
	\begin{equation}\label{def:problem_instance}
		(\mathbb{B}, \mathbb{B}^{\prime\prime}, \Theta, \mathbf{y}) \in 
		\mathbb{R}^{(MK+p) \times (MK+p)} \times 
		\mathbb{R}^{(MK+p) \times (MK+p)} \times 
		\mathbb{R}^{n \times (MK+p)} \times 
		\mathbb{R}^n,
	\end{equation}
	where $B, B^{\prime\prime} \in \mathbb{R}^{M \times M}$ are the Gram matrices of the basis functions and their second derivatives, with entries
	\begin{equation}
		B_{m, m'} := \int_0^1 b_m(t)\, b_{m'}(t)\, dt, \quad
		B^{\prime\prime}_{m, m'} := \int_0^1 b_m''(t)\, b_{m'}''(t)\, dt,
	\end{equation}
	for $m, m' = 1, \ldots, M$. We then define the block-diagonal matrices
	\begin{equation}\label{def:gram_block}
		\mathbb{B} := \operatorname{blkdiag}(B, \ldots, B, I_p), \quad
		\mathbb{B}^{\prime\prime} := \operatorname{blkdiag}(B^{\prime\prime}, \ldots, B^{\prime\prime}, I_p),
	\end{equation}
	the full design matrix $\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$, and the response vector $\mathbf{y} := (y_1, \ldots, y_n)^\top$.

@@

	\subsection{Step 2: Computing Residualized Predictors via Hybrid-on-Scalar Regression} \label{subsec: Step 2: Computing Predictors via Hybrid-on-Scalar Regression}
The $l$-th iteration's second step involves orthogonalization. This is achieved by fitting a hybrid-on-scalar regression model, $$\widetilde{W}^{[l]}_i = \hat{\rho}^{[l]}\, \delta^{[l]} + E_i^{[l]},$$ and then taking the residuals. Here, $\delta^{[l]}$ and $E_i^{[l]}$ are the regression coefficient and error terms, respectively, both elements of $\tilde{\mathcal{H}}$ defined in \eqref{def:truncated_hilbert_space}. In the same spirit as in Section \ref{sec: Regularization of PLS Components}, we want the estimated regression coefficieint $\hat{\delta}^{[l]} 
= 
(\hat{\delta}^{[l]}_1, \ldots, \hat{\delta}^{[l]}_K, \hat{\boldsymbol{\chi}}^{[l]} )$ 
have  smooth functional parts, for the whole PLS estimation proecedure to be stable.
Therefore, we obtain $\delta^{[l]}$ by minimizing the followign penaluized least square criteria, where the functional part is penalized by 
 integeraed squared second derivative defined in \eqref{def:roughness_penalty}:
 \begin{align*}
\operatorname{PENSSE}( \delta)
&=
\sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta,
\,
\widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta
\rangle_{\mathcal{H}}
%%%%%%%%%%%%%%%%%%%%
\\&=
\sum_{i=1}^n 
\langle \widetilde{W}_i^{[l]},  \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}
+
\sum_{i=1}^n 
\langle \delta,  \, \delta \rangle_{\mathcal{H}}
-
\sum_{i=1}^n 
( \hat{\rho}_i^{[l]} )^2
\langle \delta,  \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}
 \end{align*}

%As in Section \ref{subsec: Step 1: Computing Smoothed PLS components}, the hybrid response can be expanded using basis functions as $\widetilde{W}[\mathbf{t}] = \widetilde{C}^* B^*[\mathbf{t}]$. Similarly, we expand each functional regression coefficient using the basis functions as $\pi^{(k)}(t_k) =\sum_{m=1}^M d_m^{(k)} b^{(k)}_m(t_k) = \boldsymbol{\gamma}^{(k)\top} \mathbf{b}^{(k)}(t_k)$, where $\boldsymbol{\gamma}^{(k)} = (d_1^{(k)}, \ldots, d_M^{(k)})\top \in \mathbb{R}^M$ are basis coefficients. Here, a different basis system can be flexibly chosen to expand the regression coefficient, but without loss of generality we use the same system as in Section \ref{subsec: Step 1: Computing Smoothed PLS components}. With the $(MK+p)$-dimensional row vector $\boldsymbol{\gamma}^* = (\boldsymbol{\gamma}^{(1)\top}, \ldots, \boldsymbol{\gamma}^{(K)\top}, \gamma_1, \ldots, \gamma_p)$ that stacks the basis coefficients and scalar regression coefficients, we can express the linear predictor of the regression as: 
$\widehat{\boldsymbol{\rho}} \, \boldsymbol{\delta}[\mathbf{t}]^\top = \widehat{\boldsymbol{\rho}} \, \boldsymbol{\gamma}^* B^*[\mathbf{t}]$. Finally, these basis expansions of the hybrid response and the regression coefficient enable us to re-write the hybrid-on-scalar regression model \eqref{eq: hybrid on scalar regression model} in a finite-dimensional form:
\begin{equation}
	\widetilde{C}^* B^*[\mathbf{t}] = \widehat{\boldsymbol{\rho}} \, \boldsymbol{\gamma}^* B^*[\mathbf{t}] + \mathbf{e}[\mathbf{t}].
	\label{eq: rewritten hybrid on scalar regression model}
\end{equation}
$\mathbf{e}[\mathbf{t}]$ is an $n \times (K+p)$ matrix of errors. 
\\

>>>>

	\subsection{Preliminary step: Finite-basis Approximation }\label{sec: finite basis approximation}	
	Let ${b_m(t)}$ be a twice-differentiable basis of $L^2([0,1])$ whose second derivatives are also linearly independent, for example, cubic B-splines, the Fourier basis, or an orthonormal polynomial basis of degree greater than three. Using this basis, 
	the $j$th functional predictor, 
	regression coefficient, PLS component direction, and orthogonalization regression coefficient (with iteration indices suppressed) are represented as follows:
		X_{ij}(t) = \sum_{m=1}^\infty \theta_{ijm} \, b_m(t),~
		\beta_j(t) = \sum_{m=1}^\infty \eta_{jm} \, b_m(t),~
		%%%%%%%%%%%%%%%%%%%%%%
		\xi_j(t)
		=
		\sum_{m=1}^\infty
		\gamma_{jm}
		b_m(t),
		~
		\delta_j(t)
=
\sum_{m=1}^\infty
\pi_{jm}
b_m(t)		
	In practice, the full set of coefficients can not be obtained  with finite sample size, as functional data are measured on a finite grid. Thus, we truncate the expansion at $M$ terms and define the truncated basis space
	\mathcal{B} := \operatorname{span}\bigl(b_1(t), \ldots, b_M(t)\bigr).
	We choose a moderately large $M$ (e.g., 15 or 20) to capture functional variation without fine-tuning, as smoothness is handled via penalization (see Section~\ref{sec: Regularization of PLS Components}).
	The truncated expansions of the predictor and coefficient are denoted as
		\widetilde{X}_{ij}(t) := \sum_{m=1}^M \theta_{ijm} \, b_m(t), \quad
		\tilde{\beta}_j(t) := \sum_{m=1}^M \eta_{jm} \, b_m(t),
	and our suggest method restricts each PLS component direction and orthogonalization regression coefficient to admit the following expansion:
	\begin{equation}\label{def:truncated_zeta_and_delta}
		\xi_j(t) = \sum_{m=1}^M \gamma_{jm} \, b_m(t),
		\quad
				\delta_j(t)
		=
		\pi_{jm}
		b_m(t)		
	This implies that all computations in this paper are carried out entirely within the subspace
	\begin{equation}\label{def:truncated_hilbert_space}
		\widetilde{ \mathcal{H} } := \mathcal{B}^K \times \mathbb{R}^p \subset \mathcal{H}.
	The $i$th hybrid predictor, projected on $\widetilde{ \mathcal{H} }$, is represented by the tuple
		\widetilde{W}_i := 
		\widetilde{X}_{i1}, \ldots, \widetilde{X}_{iK}, \boldsymbol{Z}_i
		).
	Let $\boldsymbol{\theta}_{ij}$, $\boldsymbol{\eta}_j$,  $\boldsymbol{\gamma}_j$, and
	$\boldsymbol{\pi}_j$
	 denote the $M$-dimensional vectors of coefficients:
	\begin{equation}\label{def:coef_vector_form}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top,
~
		\boldsymbol{\eta}_{j} := (\eta_{j1}, \ldots, \eta_{jM})^\top ,~
		\boldsymbol{\gamma}_{j} := (\gamma_{j1}, \ldots, \gamma_{jM})^\top,~
			\boldsymbol{\pi}_{j} := (\pi_{j1}, \ldots, \pi_{jM})^\top
	For the predictors, we stack the coefficient vectors across observations into the matrix
	\begin{equation}\label{def:theta}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	The full data for the hybrid PLS problem can be represented by the tuple
	\begin{equation}\label{def:problem_instance}
		(\mathbb{B}, \mathbb{B}^{\prime\prime}, \Theta, \mathbf{y}) \in 
		\mathbb{R}^{(MK+p) \times (MK+p)} \times 
		\mathbb{R}^{(MK+p) \times (MK+p)} \times 
		\mathbb{R}^{n \times (MK+p)} \times 
		\mathbb{R}^n,
	where $B, B^{\prime\prime} \in \mathbb{R}^{M \times M}$ are the Gram matrices of the basis functions and their second derivatives, with entries
	\begin{equation}\label{def:B_matrix}
		B_{m, m'} := \int_0^1 b_m(t)\, b_{m'}(t)\, dt, \quad
		B^{\prime\prime}_{m, m'} := \int_0^1 b_m''(t)\, b_{m'}''(t)\, dt,
	for $m, m' = 1, \ldots, M$. We then define the block-diagonal matrices
	\begin{equation}\label{def:gram_block}
		\mathbb{B} := \operatorname{blkdiag}(B, \ldots, B, I_p), \quad
		\mathbb{B}^{\prime\prime} := \operatorname{blkdiag}(B^{\prime\prime}, \ldots, B^{\prime\prime}, I_p),
	the full design matrix $\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$, and the response vector $\mathbf{y} := (y_1, \ldots, y_n)^\top$.

@@

	\subsection{Step 2: Computing Residualized Predictors via Hybrid-on-Scalar Regression} \label{subsec: Step 2: Computing Predictors via Hybrid-on-Scalar Regression}
The $l$-th iteration's second step involves orthogonalization. This is achieved by fitting a hybrid-on-scalar regression model, where the predictor is the PLS score for each observation. The individual PLS score is computed as  $\hat{\rho}^{[l]}_i := \langle \widetilde{W}_i^{[l]}, \, \widehat{\xi}^{[l]} \rangle_{\mathcal{H}}$, utilizing the estimated PLS component direction $\widehat{\xi}^{[l]}$ obtained from Propositions \ref{proposition:eigen_regul} and \ref{proposition:linear_regul}.Subsequently, using the basis expansion coefficient notation from \eqref{def:truncated_zeta_and_delta}, \eqref{def:coef_vector_form}, \eqref{def:theta}, and \eqref{def:B_matrix}, the full vector of PLS scores $\hat{\boldsymbol{\rho}}^{[l]} := (\hat{\rho}^{[l]}_1, \ldots, \hat{\rho}^{[l]}_n)^\top$ is computed through the following matrix multiplication:
\begin{equation*}
( \hat{\boldsymbol{\rho}}^{[l]} )^\top = \sum_{k=1}^K ( \hat{\boldsymbol{\gamma}}_j^{[l]})^\top B \, (\Theta_j^{[l]})^\top + \boldsymbol{\zeta}^\top (\mathbf{Z}^{[l]}).
\end{equation*}
Using this as a scalar predictor vector, we fit the regression model $\widetilde{W}^{[l]}_i = \hat{\rho}^{[l]}_i\, \delta^{[l]} + E_i^{[l]}$, and then take the residuals as the predictor for the next iteration, namely $\widetilde{W}^{[l+1]}_i$. Here, $\delta^{[l]}$ and $E_i^{[l]}$ are the regression coefficient and error terms, respectively, both elements of $\tilde{\mathcal{H}}$ as defined in \eqref{def:truncated_hilbert_space}. To ensure the stability of the entire PLS estimation procedure, the estimated regression coefficient $\hat{\delta}^{[l]} = (\hat{\delta}^{[l]}_1, \ldots, \hat{\delta}^{[l]}_K, \hat{\boldsymbol{\chi}}^{[l]} )$ should have smooth functional parts, consistent with Section \ref{sec: Regularization of PLS Components}. Consequently, $\delta^{[l]}$ is obtained by minimizing a penalized least squares criterion, where the functional part is penalized by the integrated squared second derivative defined in \eqref{def:roughness_penalty}.
 \begin{align*}
\operatorname{PENSSE}( \delta)
&=
\sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta,
\,
\widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta
\rangle_{\mathcal{H}}
+
\tau
\sum_{j=1}^K\operatorname{PEN}(\delta_j) 
%%%%%%%%%%%%%%%%%%%%
\\&=
\sum_{i=1}^n 
\langle \widetilde{W}_i^{[l]},  \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}
-
\sum_{i=1}^n 
 \hat{\rho}_i^{[l]} 
\langle \delta,  \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}
+
\bigl(
\sum_{i=1}^n 
( \hat{\rho}_i^{[l]} )^2
\bigr)
\langle \delta,  \, \delta \rangle_{\mathcal{H}}
+
\tau
\sum_{j=1}^K\operatorname{PEN}(\delta_j) 
\\&=
 \operatorname{tr}(C^{*T} C^{*}J^{*}\Lambda) 
 - 
 2 \operatorname{tr}(d^{*}J^{*}C^{*T} \rho\Lambda) 
 +
 \rho^T \rho
  \operatorname{tr}( d^{*}J^{*}d^{*T} \Lambda) + \tau \operatorname{tr}(\ddot{d}^{*}J^{*}\ddot{d}^{*T}) \jongmin{need to be fixed}
 \end{align*}

%Using the basis expansion coefficient notation presented in \eqref{def:truncated_zeta_and_delta} and \eqref{def:coef_vector_form}, \eqref{def:theta}, \eqref{def:B_matrix}
\jongmin{Old writing starts here. I temporarily mixed the order of the paragraphs.}
%As in Section \ref{subsec: Step 1: Computing Smoothed PLS components}, the hybrid response can be expanded using basis functions as $\widetilde{W}[\mathbf{t}] = \widetilde{C}^* B^*[\mathbf{t}]$. Similarly, we expand each functional regression coefficient using the basis functions as $\pi^{(k)}(t_k) =\sum_{m=1}^M d_m^{(k)} b^{(k)}_m(t_k) = \boldsymbol{\gamma}^{(k)\top} \mathbf{b}^{(k)}(t_k)$, where $\boldsymbol{\gamma}^{(k)} = (d_1^{(k)}, \ldots, d_M^{(k)})\top \in \mathbb{R}^M$ are basis coefficients. Here, a different basis system can be flexibly chosen to expand the regression coefficient, but without loss of generality we use the same system as in Section \ref{subsec: Step 1: Computing Smoothed PLS components}. With the $(MK+p)$-dimensional row vector $\boldsymbol{\gamma}^* = (\boldsymbol{\gamma}^{(1)\top}, \ldots, \boldsymbol{\gamma}^{(K)\top}, \gamma_1, \ldots, \gamma_p)$ that stacks the basis coefficients and scalar regression coefficients, we can express the linear predictor of the regression as: 
$\widehat{\boldsymbol{\rho}} \, \boldsymbol{\delta}[\mathbf{t}]^\top = \widehat{\boldsymbol{\rho}} \, \boldsymbol{\gamma}^* B^*[\mathbf{t}]$. Finally, these basis expansions of the hybrid response and the regression coefficient enable us to re-write the hybrid-on-scalar regression model \eqref{eq: hybrid on scalar regression model} in a finite-dimensional form:
\begin{equation}
	\widetilde{C}^* B^*[\mathbf{t}] = \widehat{\boldsymbol{\rho}} \, \boldsymbol{\gamma}^* B^*[\mathbf{t}] + \mathbf{e}[\mathbf{t}].
	\label{eq: rewritten hybrid on scalar regression model}
\end{equation}
$\mathbf{e}[\mathbf{t}]$ is an $n \times (K+p)$ matrix of errors. 
\\
